---
title: "Homework 5"
author: "Nikhita Gandhe"
date: "04/22/2025"
format: pdf
---

Please find the link to my GitHub repository here: https://github.com/nsgand2/ECON-470-HWK5.git 

# Question 1
Plot the share of the adult population with direct purchase health insurance over time.

```{r, echo=FALSE, warning=FALSE, message=FALSE}


if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, data.table, gdata)

# Load Data
df <- read.table("C:/Users/Nikhita Gandhe/Documents/GitHub/ECON-470-HWK5/data/output/acs_medicaid.txt", header=TRUE, sep = "\t")


# Inspect your column names
names(df)

# Summarize by year
dp_by_year <- df %>%
  # drop missing values
  filter(!is.na(ins_direct)) %>%    
  mutate(dir_share = ins_direct/adult_pop) %>%
  group_by(year) %>%                     
  summarise(
    avg_direct_purchase = mean(ins_direct, na.rm = TRUE)
  )


# Plot it
library(ggplot2)
ggplot(dp_by_year, aes(x = year, y = avg_direct_purchase)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(dp_by_year$year)) +
  labs(
    title = "Share of Adult Population with Direct-Purchase Health Insurance",
    x     = "Year",
    y     = "Direct-Purchase Share (%)"
  ) +
  theme_minimal()
```

# Question 2
Discuss the reduction in direct purchase health insurance in later years. Can you list a couple of policies that might have affected the success of the direct purchase insurance market?


 In the later years (especially post-2016), there was a noticeable decline in direct-purchase health insurance. This decline can be attributed to several key policy changes:

 1. Repeal of the Individual Mandate Penalty (2017 Tax Cuts and Jobs Act)
 This penalty was a core component of the Affordable Care Act (ACA).

 Repealing it reduced the incentive for healthy individuals to purchase insurance, which led to adverse selection in the individual market (higher premiums, fewer enrollees).

 2. Expansion of Short-Term and Association Health Plans (Trump Admin, 2018)
 These alternatives were often cheaper but offered fewer benefits and protections.

 They siphoned younger, healthier consumers out of ACA-compliant plans, destabilizing the risk pool in the individual market.

 3. Cuts to ACA Outreach and Enrollment Assistance
 Federal support for advertising and enrollment assistance was slashed between 2017–2019.

 As a result, fewer people signed up for Marketplace plans, contributing to a decline in coverage via direct purchase.

# Question 3
Plot the share of the adult population with Medicaid over time.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Summarize Medicaid share by year
medicaid_by_year <- df %>%
  
  mutate(medicaid_share = ins_medicaid / adult_pop) %>%
  group_by(year) %>%
  summarise(
    avg_medicaid_share = mean(medicaid_share, na.rm = TRUE)
  )

# Plot it
ggplot(medicaid_by_year, aes(x = year, y = avg_medicaid_share)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(medicaid_by_year$year)) +
  labs(
    title = "Share of Adult Population with Medicaid Coverage",
    x = "Year",
    y = "Medicaid Coverage Share"
  ) +
  theme_minimal()
```

# Question 4
# Plot the share of uninsured over time, separately by states that expanded Medicaid in 2014 versus those that did not. Drop all states that expanded after 2014.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Filter to only states that either expanded in 2014 or never expanded (drop post-2014 expanders)
q4_data <- df %>% 
  filter(expand_year == 2014 | is.na(expand_year)) %>%  # drop states that expanded after 2014
  mutate(
    expand_2014 = ifelse(expand_year == 2014, 1, 0),     # indicator: 1 = expanded in 2014, 0 = never expanded
    unins = uninsured / adult_pop                       # calculate uninsured share
  ) %>%
  group_by(expand_2014, year) %>%
  summarise(avg_unins = mean(unins, na.rm = TRUE)) %>%
  ungroup()

# Plot
q4_plot <- ggplot(q4_data, aes(x = year, y = avg_unins, color = factor(expand_2014), group = factor(expand_2014))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Average Uninsured Share by Medicaid Expansion Status (2014)",
    x = "Year",
    y = "Average Uninsured Share",
    color = "Expanded in 2014"
  ) +
  scale_color_manual(values = c("0" = "red", "1" = "blue"),
                     labels = c("0" = "Never Expanded", "1" = "Expanded in 2014")) +
  theme_minimal()

# Print plot
q4_plot

```

# Question 5
Calculate the average percent of uninsured individuals in 2012 and 2015, separately for expansion and non-expansion states. Present your results in a basic 2x2 DD table.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Step 1: Calculate uninsurance rate for 2012 and 2015, by expansion group
dd_table <- df %>%
  filter(year %in% c(2012, 2015), expand_year == 2014 | is.na(expand_year)) %>%
  mutate(
    expand_2014 = ifelse(expand_year == 2014, 1, 0),
    unins = uninsured / adult_pop
  ) %>%
  group_by(expand_2014, year) %>%
  summarise(avg_unins = mean(unins, na.rm = TRUE)) %>%
  ungroup()

# Step 2: Reshape to make a 2x2 table (pre vs post period)
dd_table_wide <- dd_table %>%
  pivot_wider(names_from = year, values_from = avg_unins) %>%
  rename(`Expansion Status` = expand_2014, `2012` = `2012`, `2015` = `2015`)

# Step 3: Clean up labels
dd_table_wide$`Expansion Status` <- ifelse(dd_table_wide$`Expansion Status` == 1, "Expanded in 2014", "Never Expanded")

# Step 4: Display table (in RMarkdown or notebook)
knitr::kable(dd_table_wide, format = "html", caption = "Average Uninsured Rate by Expansion Status (2012 vs 2015)")

```

# Question 6
Estimate the effect of Medicaid expansion on the uninsurance rate using a standard DD regression estimator, again focusing only on states that expanded in 2014 versus those that never expanded.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Load necessary packages if not already loaded
pacman::p_load(dplyr, modelsummary)

# Step 1: Filter to only 2014 expanders or never-expanders
reg.df <- df %>%
  filter(expand_year == 2014 | is.na(expand_year), !is.na(expand_ever)) %>%
  mutate(
    perc_unins = uninsured / adult_pop,
    post = ifelse(year >= 2014, 1, 0),
    treat = post * expand_ever  # interaction term
  )

# Step 2: Run Difference-in-Differences regression
dd <- lm(perc_unins ~ post + expand_ever + treat, data = reg.df)

# Step 3: Show results using modelsummary (clustered by State)
modelsummary(
  list("DD Estimate (2014 vs Never)" = dd),
  shape = term + statistic ~ model,
  gof_map = NA,
  coef_omit = 'Intercept',
  vcov = ~State,  # cluster standard errors at the state level
  stars = TRUE
)

```

# Question 7
Include state and year fixed effects in your estimates. Try using the lfe or fixest package to estimate this instead of directly including the fixed effects.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Load required package
library(fixest)

# Step 1: Create TWFE regression with state and year fixed effects
dd.fe <- feols(
  perc_unins ~ treat | State + year,   # treat is the DiD interaction term
  data = reg.df, 
  cluster = ~State                     # cluster standard errors at the state level
)

# Step 2: Compare simple DiD vs TWFE
msummary(
  list("Simple DD" = dd, "TWFE DD" = dd.fe),
  shape = term + statistic ~ model,
  gof_map = NA,
  coef_omit = 'Intercept',
  stars = TRUE
)

```

# Question 8
Repeat the analysis in question 7 but include all states (even those that expanded after 2014). Are your results different? If so, why?

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Step 1: Use all states, including those that expanded after 2014
reg.df2 <- df %>% 
  mutate(
    perc_unins = uninsured / adult_pop,
    post = ifelse(year >= 2014, 1, 0),
    treat = post * expand_ever   # now includes all expansion states regardless of timing
  )

# Step 2: Run basic DD regression
dd2 <- lm(perc_unins ~ post + expand_ever + treat, data = reg.df2)

# Step 3: Run TWFE regression with state & year fixed effects
dd.fe2 <- feols(perc_unins ~ treat | State + year, data = reg.df2, cluster = ~State)

# Step 4: Compare all models
msummary(
  list(
    "DD (2014 only)"     = dd,
    "TWFE (2014 only)"   = dd.fe,
    "DD (All States)"    = dd2,
    "TWFE (All States)"  = dd.fe2
  ),
  shape = term + statistic ~ model,
  gof_map = NA,
  coef_omit = 'Intercept',
  stars = TRUE
)

```


# Question 9
Provide an “event study” graph showing the effects of Medicaid expansion in each year. Use the specification that includes state and year fixed effects, limited to states that expanded in 2014 or never expanded.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

mod.twfe <- feols(perc_unins~i(year, expand_ever, ref=2013) | State + year,
                  cluster=~State,
                  data=reg.df)

iplot(mod.twfe, 
      xlab = 'Time to treatment',
      main = 'Event study')

```

# Question 10
Repeat part 9 but again include states that expanded after 2014. Note: this is tricky…you need to put all states onto “event time” to create this graph.

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Load required libraries
pacman::p_load(fixest, ggplot2, dplyr)

# Step 1: Create event time variable
reg.dfall <- df %>%
  filter(!is.na(expand_year)) %>%  # Only keep expansion states (for event time)
  mutate(
    perc_unins = uninsured / adult_pop,
    event_time = year - expand_year
  )

# Step 2: TWFE Event-Study Regression using `feols`
# This will estimate dynamic treatment effects relative to expansion
mod.twfe <- feols(
  perc_unins ~ i(event_time, ref = -1) | State + year,
  cluster = ~State,
  data = reg.dfall
)

# Step 3: Plot using built-in `iplot` from fixest
iplot(
  mod.twfe,
  xlab = "Years from Medicaid Expansion",
  main = "Event Study: Uninsured Rate Response to Medicaid Expansion"
)

```